\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Yan's edits

\linenumbers
\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high level, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely,ignatieva2023distribution}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a stochastic process or of an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix XXX).

% JK: this para need a major update, it's quite weak right now.
The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank},
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks for the phrase Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their constituent genomes and paths
of genetic inheritance.
We label nodes using using lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
positions and associated intervals.
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombinations that gave rise to these genomes.
% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.

@@

Outside of the context of simulation, ARG-unary nodes are
only like to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-Cov-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Locally unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain locally unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore,
several current ARG inference methods infer locally unary nodes (see section XXX).

@@

>>>>

\linenumbers
\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely,ignatieva2023distribution}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a stochastic process or a
genetic genealogy inferred from a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix XXX).

% JK: this para need a major update, it's quite weak right now.
The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank},
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined, which causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, definitions linked to
the original stochastic process lead to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks for the phrase Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
using using lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombinations that gave rise to these genomes.
% Following the above definitions, corresponding data for this gARG is shown in Supplementary? Table XXX.

@@

Outside of the context of simulation, ARG-unary nodes are
only like to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-Cov-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore,
several current ARG inference methods infer locally unary nodes (see section XXX).

@@

\section{Additional plots}

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
	\end{center}
	\caption{\label{fig-simplification-with-edges}
	The four levels of ARG simplification from Fig.~\ref{fig-simplification}A--D, with edge intervals shown.
	}
\end{figure}
